---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r echo=FALSE, message=FALSE}

library(tidyverse)
data<-read_csv("~/Downloads/New Compressed (zipped) Folder/A.csv")

data_clean<-data%>%janitor::clean_names()
```

## 1.3. Task
Summary stats

```{r echo=FALSE}
data_clean%>% skimr::skim()
```

Refund times:

```{r echo=FALSE}
max(data_clean$refund_time, na.rm=TRUE)

data_clean%>%group_by(satisfaction_rating)%>%
                      summarise(average_refund_time=mean(refund_time, na.rm=TRUE))

data_clean%>%count(refund_time, sort=TRUE)


data_clean$time_between

data_clean%>%
  group_by(refund_time)%>%
  summarise(average_satisfaction=mean(satisfaction_rating, na.rm=TRUE))%>%
  arrange(refund_time)

data_clean%>%
  group_by(booking_channel)%>%
  summarise(average_satisfaction=mean(satisfaction_rating, na.rm=TRUE),
            average_refund_time=mean(refund_time, na.rm=TRUE))

data_clean%>%
  group_by(traveling_for_work)%>%
  summarise(average_satisfaction=mean(satisfaction_rating, na.rm=TRUE),
            average_refund_time=mean(refund_time, na.rm=TRUE))
  
data_clean%>%
  group_by(card_type)%>%
  summarise(average_refund_time=mean(refund_time, na.rm=TRUE))%>%
  arrange(desc(average_refund_time))

```

10 days in refund time is a clear outlier with only two cases

These are some boxplots if you want to include them

```{r echo=FALSE}
boxplot(data_clean$refund_time, ylab="Refund time", col="bisque")

boxplot(satisfaction_rating~refund_time, data=data_clean, col="bisque", xlab="Refund time", ylab="Satisfaction rating")
```

To explore relationship you can also use regression (=linear regression)

Correlation

```{r echo=FALSE}
data_corr<-data_clean%>%
  select(satisfaction_rating, refund_time, data_privacy_concerns, brand_loyalty)
  
data_corr<-data_corr[rowSums(is.na(data_corr)) != 1, ]



cormat <- round(cor(data_corr),2)
cormat
ggcorrplot::ggcorrplot(cormat)
```

If it is closer to 1 - it is a strong relationship, e.g. there is a strong negative relationship between satisfaction ratings and refund time. The relationship between data privacy and satisfaction rating is negative but much weaker (not a serious issue...). Data privacy does have positive relationship with refund time, but not really strong (less than 0.5)

Brand loyalty and satisfaction are highly correlated, such customers do not have data privacy concerns

```{r echo=FALSE} 
ggplot(data_clean, aes(x = refund_time, y = satisfaction_rating)) + 
  geom_point() + stat_smooth(method = "lm", col="red")

```


