---
title: "Longitudinal invariance"
author: "Maria"
format: html
editor: visual
---

## Setup

```{r}
library(semTools)
library(haven)
library(semPlot)
library(lavaan)
library (psych)
library (psychTools)
library(tidyverse)
library (esem)

```

### EFA on wave 1 

Data prep - will be hidden in the package
waves are not "hard-coded" in variable names, but included as a variable `wave`

```{r}
data_full<-read_sav("data.sav")

data_full<-data_full[,1:76]


#remove spss column names
data_full <- zap_formats(zap_labels(data_full))

data_full<-data_full%>%pivot_longer(
  cols=!hicid,
  names_to = c("var", "wave"),
  names_sep = "_",
  values_to = "score"
)%>%
  pivot_wider(
    names_from=var,
    values_from=score
  )
```

EFA is done on wave 1 to generate the model syntax that will be tested further in the second stage which is CFA

```{r}

colnames(data_full)

data_wave1<-data_full%>%
  filter(wave==1)%>%
  select(-hicid, -wave)%>%
  drop_na()
```


## ESEM fitting 

Making target matrix `target_wave1` with wave 1 data and 5 factors, specified in `main_loadings_list`

```{r}
main_loadings_list <- list(
  pp = c("s6", "s11", "s14", "s19", "s23"),
  cp = c("s5", "s7", "s12", "s18", "s22"),
  es = c("s3", "s8", "s13", "s16", "s24"),
  ha = c("s2","s10","s15","s21","s25"),
  ps = c("s1","s4","s9","s17","s20")
)

target_wave1<-make_target(
  data=data_wave1, 
  keys=main_loadings_list)
```

EFA with for wave 1 using rotation matrix `target_wave1`

```{r}
esem_efa_results_wave1<-esem_efa(
  data=data_wave1,
  nfactors = 5,
  rotate="TargetQ",
  Target= target_wave1)


esem_model_wave1 <- esem_syntax(esem_efa_results_wave1, referent_list)
```

Produced the following syntax to be used for testing

```{r}
writeLines(esem_model_wave1)
```
The model (=syntax) does not include wave number in the name of the variable, making this syntax an easy way to apply to all wave for longitudional invariance.

### CFA - longitudinal invariance models

Longitudinal (or multiple-group) CFAs are done using `cfa()` in `lavaan` with additional parameters (`group`, `group.equal`, and `group.partial`). These parameters specify equality constraints between the different waves (=groups).

The models can also be specified manually and compared using `anova()` function.

```{r}

data_wave1<-data_full%>%
  filter(wave==1)%>%
  select(-hicid, -wave)%>%
  drop_na()

config <- cfa(esem_model_wave1, 
              data=data_full, 
              estimator = "WLSMV",
              group="wave") 

weak <- cfa(esem_model_wave1,
            data=data_full, 
            group="wave",
            estimator = "WLSMV",
            group.equal="loadings")

strong<- cfa(esem_model_wave1,
             data=data_full, 
             group="wave",
            estimator = "WLSMV",
             group.equal = c("loadings", "intercepts"))


strict<- cfa(esem_model_wave1,
             data=data_full, 
             group="wave", 
             estimator = "WLSMV",
             group.equal = c("loadings", "intercepts","residuals"))

anova(config, weak, strong, strict) 

```
### Modification indices

#### Configural model

```{r}
modificationindices(config)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(config) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Weak model

```{r}
modificationindices(weak)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(weak) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Strong model

```{r}
modificationindices(strong)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(strong) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Strict model

```{r}
modificationindices(strict)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(strict) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

### MLR

```{r}
config <- cfa(esem_model_wave1, 
              data=data_full, 
              estimator = "MLR",
              group="wave") 

weak <- cfa(esem_model_wave1,
            data=data_full, 
            group="wave",
            estimator = "MLR",
            group.equal="loadings")

strong<- cfa(esem_model_wave1,
             data=data_full, 
             group="wave",
            estimator = "MLR",
             group.equal = c("loadings", "intercepts"))


strict<- cfa(esem_model_wave1,
             data=data_full, 
             group="wave", 
             estimator = "MLR",
             group.equal = c("loadings", "intercepts","residuals"))

anova(config, weak, strong, strict) 

```

### Modification indices for MLR

#### Configural model

```{r}
modificationindices(config)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(config) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Weak model

```{r}
modificationindices(weak)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(weak) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Strong model

```{r}
modificationindices(strong)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(strong) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```

#### Strict model

```{r}
modificationindices(strict)%>%
   knitr::kable(caption="MI for all 3 time frames")


modificationindices(strict) %>%
  as_data_frame() %>%
  arrange(-mi) %>%
  filter(mi > 11) %>%
  select(lhs, op, rhs, mi, epc) %>%
  knitr::kable(caption="Largest MI values for esem_cfa_results_all", digits=3)
```
