---
title: "Sample size calculation for SDQ LSAC"
format: html
editor: visual
---

## Sample size

Planning for the sample size is essential in ESEM. This is critical to avoid issues with convergence of the models. 

With the ESEM, estimation of the sample size is suggested to be done considering traditional approaches for SEM, such as most popular estimation based on the following:

- the non-centrality parameter (Satorra and Saris, 1984)

- the model's potential to obtain an acceptable RMSEA value (MacCallum et al., 1996), or 

- Monte Carlo simulations (Muthén and Muthé). The latter is more common due to its availability in Mplus. 

Monte Carlo simulations estimation is available in R in the `simsem` package. While estimation via RMSEA can be done using `semPower` package.


## Monte Carlo simulations estimation

This estimation uses the previously develobed CFA model and factor structure of the initial instrument:

```{r}
library(esem)
library(tidyverse)
library(lavaan)
library(semPlot)
library(psych)


esem_efa_results <- esem_efa(data=sdq_lsac, 
                      nfactors =5,
                      fm = 'ML',
                      rotate="geominT", 
                      scores="regression", 
                      residuals=TRUE, 
                      missing=TRUE)

esem_model <- esem_syntax(esem_efa_results)

writeLines(esem_model)
```


A Monte Carlo simulation can be implemented by the `sim` function:

```{r}
library(simsem)
analyzeModel<-"
ML1 =~ s6_1+s11_1R+s14_1R+s19_1+s23_1
ML2 =~  s5_1+s7_1R+s12_1+s18_1+s22_1
ML3 =~  s3_1+s8_1+s13_1+s16_1+s24_1
ML4 =~  s2_1+s10_1+s15_1+s21_1R+s25_1R
ML5 =~  s1_1+s4_1+s9_1+s17_1+s20_1
"

Output <- sim(1000, analyzeModel, n=3836, generate=esem_model, lavaanfun = "cfa")

summary(Output)
```

## Estimation via RMSEA

Obtain the df of a model provided as lavaan model string (this requires the lavaan package):

```{r}
library(semPower)

semPower.getDf(esem_model)
```


Determine the achieved power with a sample size of N = 3836 to detect misspecifications of a model (involving df = 190 degrees of freedom) corresponding to RMSEA = .05 on an alpha error of .05:

```{r}
ph <- semPower.postHoc(effect = .05, effect.measure = 'RMSEA',
                         alpha = .05, N = 3836, df = 190)
summary(ph)
```


Determine the critical chi-square such that the associated alpha and beta errors are equal, assuming sample size of N = 1000, a model involving df = 100 degrees of freedom, and misspecifications corresponding to RMSEA = .05:

```{r}
cp <- semPower.compromise(effect = .05, effect.measure = 'RMSEA',
                            abratio = 1, N = 3836, df = 190)
summary(cp)
```

Plot power as function of the sample size to detect misspecifications corresponding to RMSEA = .05 (assuming df = 190) on alpha = .05:

```{r}
semPower.powerPlot.byN(effect = .05, effect.measure = 'RMSEA',
                         alpha = .05, df = 190, power.min = .05, power.max = .99)
```


Plot power as function of the magnitude of effect (measured through the RMSEA assuming df = 100) at N = 500 on alpha = .05:

```{r}
semPower.powerPlot.byEffect(effect.measure = 'RMSEA', alpha = .05, N = 500,
                              df = 190, effect.min = .001, effect.max = .10)

```


----------------------
### References:

MacCallum RC, Browne MW, Sugawara HM. Power analysis and determination of sample size for covariance structure modeling. Psychol Methods. (1996) 1:130–49. doi: 10.1037/1082-989X.1.2.130

Muthén LK, Muthén BO. How to use a Monte Carlo study to decide on sample size and determine power. Struct Equ Model. (2002) 9:599–620. doi: 10.1207/S15328007SEM0904_8

Satorra A, Saris WE. Power of the likelihood ratio test in covariance structure analysis. Psychometrika. (1985) 50:83–90. doi: 10.1007/BF02294150
